---
title: "Comparing a subject against normative gait bands"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing a subject against normative gait bands}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

PhysioGaitNorm bundles **real** normative gait reference data: mean and
standard-deviation kinematic waveform bands over the gait cycle for the nine
Gait Deviation Index / Gait Profile Score variables, derived from the public
WBDS dataset (Fukuchi, Fukuchi & Duarte 2018; 24 young healthy adults,
overground comfortable-speed walking). This vignette loads the bundled norms and
compares a **synthetic** subject against them --- everything runs offline from
the installed package data.

```{r setup}
library(PhysioGaitNorm)
```

## The bundled datasets

`listGaitNorms()` returns the provenance / version manifest.

```{r catalog}
listGaitNorms()[, c("dataset", "version", "population", "n_subjects", "speed")]
```

`loadGaitNorm()` loads the mean/SD bands. Each is a `variables x cycle-points`
matrix (rows named by kinematic variable, columns by cycle percentage).

```{r load}
norm <- loadGaitNorm("adult_reference", cycle = 101)
norm
norm$variables
```

## Comparing a synthetic subject to the knee-flexion band

We build a synthetic subject by taking the normative mean knee-flexion waveform
and reducing its swing-phase peak --- a "stiff-knee" pattern --- then score each
cycle point as a z-score against the normative mean and SD.

```{r subject}
pct   <- norm$percent
mu    <- norm$mean["knee_flexion", ]
sdev  <- norm$sd["knee_flexion", ]

# synthetic subject: normative mean minus a swing-phase bump (degrees)
subject <- mu - 18 * exp(-((pct - 72)^2) / (2 * 9^2))

z <- (subject - mu) / sdev
c(max_abs_z = round(max(abs(z)), 2),
  at_percent = pct[which.max(abs(z))])
```

The largest deviation falls in swing phase, as expected. A plot makes the
subject's departure from the normal band explicit.

```{r plot}
plot(pct, mu, type = "l", lwd = 2, ylim = range(mu - 2 * sdev, mu + 2 * sdev, subject),
     xlab = "gait cycle (%)", ylab = sprintf("knee flexion (%s)", norm$units),
     main = "Subject vs normative band")
polygon(c(pct, rev(pct)), c(mu + 2 * sdev, rev(mu - 2 * sdev)),
        col = "grey88", border = NA)
lines(pct, mu, lwd = 2)
lines(pct, subject, col = "firebrick", lwd = 2)
legend("topright", c("norm mean", "+/- 2 SD", "subject"),
       col = c("black", "grey88", "firebrick"), lwd = c(2, 8, 2), bty = "n")
```

## Where next

The same `norm$mean` / `norm$sd` bands drive the Gait Deviation Index and Gait
Profile Score calculations elsewhere in the ecosystem; the per-subject feature
matrix (`loadGaitNorm(features = TRUE)$features`) is the normative population
used to build the GDI basis. See `?loadGaitNorm` and the provenance manifest in
`listGaitNorms()` for full data lineage and citations.
